---
title: "Full Untargeted Analysis of Tarragona Samples"
subtitle: "PLS-DA model with GC-IMS Data"
author: "Tecla Duran Fort"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
---

# Set Up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ggplot2)
library(pls)
library(plsVarSel)
library(dplyr)
library(pROC)
library(tidyverse)
library(purrr)
library(rprojroot)
knitr::opts_knit$set(root.dir = rprojroot::find_root(rprojroot::is_git_root))
source("../../../load_tools.R")
```

```{r, echo=FALSE}
# PARALLELISATION 

library(BiocParallel)

# Detecta nombre total de cores disponibles
n_total <- parallel::detectCores()

# Deixa 4 cores lliures per seguretat
n_workers <- max(1, n_total - 4)

message("Detected cores: ", n_total)
message("Using workers:  ", n_workers)

# Linux/macOS → Multicore
if (.Platform$OS.type == "unix") {
    param <- MulticoreParam(workers = n_workers)
} else {
    # Windows → Snow
    param <- SnowParam(workers = n_workers, type = "SOCK")
}

register(param)

```

\newpage

# Load Data

## Loading Full GCIMS Peak table 

The loaded dataset contains only the **patient samples** and has already been [filtered based on cluster representation and RSD](https://github.com/tecladuran/targetml-workflows/blob/main/docs/measurements_tarragona/cluster_filtering_summary.md)


```{r}
data <- read_csv("data/tables/tgn_gcims/peak_table_filtered_all_samples.csv")
data_patients <- data%>%filter(matrix=="urine")%>%dplyr::select(-matrix,-control_level,-day)
creatinine<- read_csv("data/tables/tgn/creatinine.csv") 
cluster_cols <- grep("^Cluster", names(data), value = TRUE)
```

# Preprocessing

## PQN Normalization

### Reference spectrum computation

A reference spectrum is required for PQN normalization.  
Here, the **median spectrum of all control samples** is used as reference.  

The **target analytes** (Anisole, 2-Heptanone (Monomer, Dimer), 2-Pentanone (Monomer, Dimer)) are intentionally excluded because the QC samples include **different concentration levels of these compounds**.  
The identification of these target analytes is detailed in the following document:  
[Identification of Target Analytes – Tarragona Urine Measurements](https://github.com/tecladuran/targetml-workflows/blob/main/docs/measurements_tarragona/preprocessing/identification/identification_tgn_urine.md)


```{r}
# Clusters corresponding to target analytes
excluded_clusters <- c(3, 155, 310, 39, 236)
excluded_names <- paste0("Cluster", excluded_clusters)

# Compute the reference (median) spectrum using control samples only
ref_spectrum <- data %>%
  filter(class == "control") %>%
  select(starts_with("Cluster")) %>%
  select(-any_of(excluded_names)) %>%
  summarise(across(everything(), median, na.rm = TRUE)) %>%
  as.numeric()

# Assign column names to the numeric vector for later reference
names(ref_spectrum) <- colnames(
  data %>%
    select(starts_with("Cluster")) %>%
    select(-any_of(excluded_names))
)

```

```{r plot-spectrum, echo=FALSE}
df_plot <- data.frame(
  index = seq_along(ref_spectrum),
  cluster  = rownames(as.data.frame(ref_spectrum)),
  value = ref_spectrum
)

ggplot(df_plot, aes(x = index, y = value)) +
  geom_line(linewidth = 0.2) +
  geom_point(size=0.5) +
  ggtitle("PQN Reference Spectrum") +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
  )


```

### Normalization 

The **Probabilistic Quotient Normalization (PQN)** corrects for dilution or total intensity differences between samples.
For each sample:

1. The quotient between its spectrum and the reference spectrum is computed.
2. The **median of the quotients** gives the normalization factor *f*.
3. All intensities in that sample are divided by *f*.

```{r}
data_pqn <- data %>%
  mutate(across(starts_with("Cluster"), as.numeric)) %>%
  group_split(row_number()) %>%                     # Split into individual samples
  map_dfr(function(row_df) {
    # Extract numeric cluster values for the current sample
    x <- as.numeric(row_df %>%
                      select(starts_with("Cluster")) %>%
                      select(-any_of(excluded_names)))
    
    # Compute the ratio between sample and reference spectrum
    quotients <- x / ref_spectrum
    quotients <- quotients[is.finite(quotients)]    # Remove non-finite values
    
    # Median of quotients = normalization factor
    f <- median(quotients, na.rm = TRUE)
    
    # Apply normalization and store the factor
    row_df %>%
      mutate(across(starts_with("Cluster"), ~ .x / f),
             PQN_factor = f)
  })
```

### PQN Factor Distribution

To check the normalization behaviour, the distribution of PQN factors across classes is visualized.

```{r plot-factor, echo=FALSE}
ggplot(data_pqn, aes(x = class, y = PQN_factor, fill = class)) +
  geom_boxplot(alpha = 0.8, color = "grey30") +
  geom_jitter(width = 0.15, alpha = 0.4) +
  labs(
    title = "PQN normalization factors by sample class",
    x = "Class",
    y = "Normalization factor"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```
```{r df, echo=FALSE}
df_norm <- data_pqn%>%filter(matrix=="urine")%>%dplyr::select(-matrix,-control_level,-day,-PQN_factor, -`row_number()`)
```


## Log Transformation 

```{r, echo=TRUE}
# ---- LOG-TRANSFORM ----
data_log <- df_norm
data_log[cluster_cols] <- log1p(data_log[cluster_cols])
```


# Exploratory Analysis


```{r}
# Simple PCA score plot function
plot_pca <- function(data, labels, levels, colors) {
  
  # Select only cluster columns (numerical features)
  features <- data[, grepl("^Cluster", colnames(data))]
  
  # Run PCA
  pca_res <- prcomp(features, scale. = TRUE)
  
  # Build score dataframe
  scores <- as.data.frame(pca_res$x[, 1:2])   # PC1 + PC2
  scores$label <- factor(labels, levels = levels)
  
  # Simple shapes for the two groups
  shapes <- c(16, 17)   # circle, triangle
  
  # Score plot
  ggplot(scores, aes(PC1, PC2, color = label, shape = label)) +
    geom_point(size = 3, alpha = 0.8) +
    scale_color_manual(values = colors) +
    scale_shape_manual(values = shapes) +
    theme_minimal(base_size = 14) +
    labs(title = "PCA Score Plot", x = "PC1", y = "PC2")
}

```

```{r}
group_labels <- df_norm$patient_condition
group_levels <- c("CTRL", "CRC")
group_colors <- c("CTRL" = "#337B9F", "CRC" = "#C83342")

p <- plot_pca(data_log, group_labels, group_levels, group_colors)

print(p)
```

# Classification Performance

## Functions Definition

### Make Stratified Folds

```{r}
make_folds <- function(..., k) {
  set.seed(1)
  folds <- vector("list", k)
  
  # Capture all input variables as a data frame
  vars <- list(...)
  if (length(vars) == 0) stop("You must provide at least one variable for stratification.")
  
  # Combine all variables into one data frame
  df <- as.data.frame(vars)
  
  # Create stratification factor (interaction between all variables)
  strata <- interaction(df, drop = TRUE)
  
  # Create folds maintaining proportional representation of each stratum
  for (lvl in levels(strata)) {
    idx <- which(strata == lvl)
    idx <- sample(idx)
    split_idx <- cut(seq_along(idx), breaks = k, labels = FALSE)
    for (i in 1:k) {
      folds[[i]] <- c(folds[[i]], idx[split_idx == i])
    }
  }
  
  return(folds)
}
```

### Nested Cross Validation
```{r}
plsda_nested_cv_rfe <- function(X, y, positive_class, negative_class, 
                                ncomp_max = 10, outer_folds = 5, inner_folds = 4) {
  
  y <- factor(y, levels = c(negative_class, positive_class))
  y_num <- ifelse(y == positive_class, 1, 0)
  
  outer_folds_idx <- make_folds(y, k=outer_folds)
  
  # --- PARALLEL OUTER LOOP ---
  results <- bplapply(seq_along(outer_folds_idx), function(o) {
    
    test_id <- outer_folds_idx[[o]]
    train_id <- setdiff(seq_along(y), test_id)
    
    X_train <- X[train_id, , drop = FALSE]
    y_train <- y[train_id]
    y_train_num <- y_num[train_id]
    X_test  <- X[test_id, , drop = FALSE]
    y_test  <- y[test_id]
    y_test_num <- y_num[test_id]
    
    features <- colnames(X_train)
    best_acc <- 0
    best_features <- features
    
    internal_acc_history <- data.frame()
    external_acc_history <- data.frame()
    
    repeat {
      inner_idx <- make_folds(y_train, k=inner_folds)
      acc_mat <- matrix(NA, nrow = inner_folds, ncol = ncomp_max)
      
      # --- INNER CV --------
      for (f in seq_along(inner_idx)) {
        val_id <- inner_idx[[f]]
        tr_id  <- setdiff(seq_along(y_train), val_id)
        
        Xi_tr <- X_train[tr_id, features, drop = FALSE]
        yi_tr <- y_train_num[tr_id]
        Xi_val <- X_train[val_id, features, drop = FALSE]
        yi_val <- y_train_num[val_id]
        
        model <- plsr(y ~ ., data = data.frame(y = yi_tr, Xi_tr),
                      ncomp = ncomp_max, method = "oscorespls", scale = TRUE)
        
        preds <- predict(model, newdata = Xi_val, ncomp = 1:ncomp_max)
        preds_mat <- matrix(preds, nrow = length(yi_val), ncol = ncomp_max)
        
        acc_mat[f, ] <- colMeans((preds_mat > 0.5) == yi_val)
      }
      
      mean_acc <- colMeans(acc_mat, na.rm = TRUE)
      opt_ncomp <- which.max(mean_acc)
      
      internal_acc_history <- rbind(
        internal_acc_history,
        data.frame(Nfeatures = length(features), Accuracy = max(mean_acc))
      )
      
      # Train full model
      model_full <- plsr(y ~ ., data = data.frame(y = y_train_num, 
                                                  X_train[, features, drop = FALSE]),
                         ncomp = opt_ncomp, method = "oscorespls", scale = TRUE)
      
      test_pred <- predict(model_full, X_test[, features, drop = FALSE], ncomp = opt_ncomp)
      test_class <- as.integer(test_pred > 0.5)
      acc_ext <- mean(test_class == y_test_num)
      
      external_acc_history <- rbind(
        external_acc_history,
        data.frame(Nfeatures = length(features), Accuracy = acc_ext)
      )
      
      if (max(mean_acc) > best_acc) {
        best_acc <- max(mean_acc)
        best_features <- features
        best_ncomp <- opt_ncomp
      }
      
      if (length(features) <= 10) break
      
      vip <- plsVarSel::VIP(model_full, opt.comp = opt_ncomp)
      drop_n <- max(1, ceiling(length(features) * 0.05))
      drop_feat <- names(sort(vip))[1:drop_n]
      features <- setdiff(features, drop_feat)
    }
    
    # Final model
    final_model <- plsr(
      y ~ ., 
      data = data.frame(y = y_train_num, X_train[, best_features, drop = FALSE]),
      ncomp = best_ncomp, method = "oscorespls", scale = TRUE
    )
    
    test_pred <- predict(final_model, X_test[, best_features, drop = FALSE], ncomp = best_ncomp)
    
    predictions <- data.frame(
      SampleIndex = test_id,
      TrueLabel = y_test_num,
      PredProb = as.numeric(test_pred),
      PredClass = as.integer(test_pred > 0.5),
      ncomp = best_ncomp
    )
    
    list(
      ncomp = best_ncomp,
      best_features = best_features,
      internal_acc_history = internal_acc_history,
      external_acc_history = external_acc_history,
      predictions = predictions
    )
  })
  
  # --- Collate predictions globally ---
  preds_all <- do.call(rbind, lapply(results, function(x) x$predictions))
  overall_acc <- mean(preds_all$TrueLabel == preds_all$PredClass)
  
  return(list(
    results = results,
    predictions = preds_all,
    overall_accuracy = overall_acc
  ))
}



```

## Evaluation


```{r}
# Predictor matrix
X <- data_log[, cluster_cols] %>% as.matrix()

# Response variable
y <- data_log$patient_condition
```


```{r}
set.seed(123)

plsda_res <- plsda_nested_cv_rfe(
  X = X,
  y = y,
  positive_class = "CRC",
  negative_class = "CTRL",
  outer_folds = 6,
  inner_folds = 5,
  ncomp_max = 10
)
```

```{r}
show_results(plsda_res, title = "PLSDA Nested CV — GC-IMS Classification")
```

